---
title: "Somatus Data Analysis Challenge"
output:
  html_document:
    theme: yeti
---

## Data Summarization Questions

```{r include=FALSE, warning = FALSE}
# Load Libraries
library(tidyverse)
library(lubridate)
library(kableExtra)
library(scales)

# Load Data
summary_data <- read_csv("./data/DE1_0_2008_Beneficiary_Summary_File_Sample_1.csv")
claims_data <- read_csv("./data/DE1_0_2008_to_2010_Inpatient_Claims_Sample_1.csv")

# Summary Data Manipulation
summary_data <- summary_data %>% 
  mutate(
    birth_date = ymd(BENE_BIRTH_DT),
    death_date = ymd(BENE_DEATH_DT),
    sex = factor(BENE_SEX_IDENT_CD, labels = c('Male','Female')),
    race = factor(BENE_RACE_CD, labels = c('White','Black','Others','Hispanic')),
    has_esrd = BENE_ESRD_IND == 'Y',
    has_alzheimer = SP_ALZHDMTA == 1,
    has_heart_failure = SP_CHF == 1,
    has_chronic_kidney = SP_CHRNKIDN == 1,
    has_cancer = SP_CNCR == 1,
    has_copd = SP_COPD == 1,
    has_depression = SP_DEPRESSN == 1,
    has_diabetes = SP_DIABETES == 1,
    has_ischemic_heart = SP_ISCHMCHT == 1,
    has_osteoporosis = SP_OSTEOPRS == 1,
    has_arthritis = SP_RA_OA == 1,
    has_stroke_tia = SP_STRKETIA == 1
)

# Inpatient Claims Data Manipulation
claims_data <- claims_data %>% 
  mutate(
    start_dt = ymd(CLM_FROM_DT),
    thru_dt = ymd(CLM_THRU_DT),
    admit_dt = ymd(CLM_ADMSN_DT),
    discharge_dt = ymd(NCH_BENE_DSCHRG_DT),
    payment_amt = CLM_PMT_AMT + NCH_PRMRY_PYR_CLM_PD_AMT
  )

claims_data <- claims_data %>% 
  unite("ICD9_str", ADMTNG_ICD9_DGNS_CD, ICD9_DGNS_CD_1:ICD9_DGNS_CD_10, sep=",", remove=FALSE, na.rm=TRUE)

# Merge Data
all_data <- summary_data %>% left_join(claims_data)

all_data <- all_data %>% 
  mutate(
    has_eskd_ckd = str_detect(ICD9_str,"5856|5855|5854|5853|5852|5851|5859") | has_esrd,
    kd_type = case_when(
      str_detect(ICD9_str,"5856") | has_esrd ~ "ESKD",
      str_detect(ICD9_str,"5855") ~ "CKD Stage 5",
      str_detect(ICD9_str,"5854") ~ "CKD Stage 4",
      str_detect(ICD9_str,"5853") ~ "CKD Stage 3",
      str_detect(ICD9_str,"5852") ~ "CKD Stage 2",
      str_detect(ICD9_str,"5851") ~ "CKD Stage 1",
      str_detect(ICD9_str,"5859") ~ "CKD Stage Unspecified",
      TRUE ~ "None"
    )
  )

all_data <- all_data %>% 
  mutate(
    kd_type = factor(kd_type, levels = c(
      "ESKD","CKD Stage 5","CKD Stage 4","CKD Stage 3","CKD Stage 2","CKD Stage 1","CKD Stage Unspecified","None","Total")
    )
  )
```


### 1. How many total CKD and ESKD members are there in the data across all years?  How many members are there by kidney disease stage?

```{r echo = FALSE, results = 'asis', message = FALSE}
grouped_counts <- all_data %>% 
  filter(kd_type != "None") %>% 
  group_by(kd_type) %>% 
  summarise(members = n_distinct(DESYNPUF_ID))

sum_row <- lapply(grouped_counts,function(z) if (is.numeric(z)) sum(z) else '')
sum_row[1] <- "Total"
results <- rbind(grouped_counts,sum_row)
colnames(results) <- c("Kidney Disease Stage","Members")

kable(results, format.args = list(big.mark = ",")) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left") %>% 
  column_spec(1, width = "15em") %>% 
  row_spec(dim(results)[1], bold = TRUE)
```

### 2. How many admissions did CKD and ESKD members have in the data across all years, by kidney disease stage?

```{r echo = FALSE, results = 'asis', message = FALSE}
grouped_counts <- all_data %>% 
  filter(kd_type != "None") %>% 
  group_by(kd_type) %>% 
  summarise(admits = n_distinct(CLM_ID))

sum_row <- lapply(grouped_counts,function(z) if (is.numeric(z)) sum(z) else '')
sum_row[1] <- "Total"
results <- rbind(grouped_counts,sum_row)
colnames(results) <- c("Kidney Disease Stage","Admissions")

kable(results, format.args = list(big.mark = ",")) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left") %>% 
  column_spec(1, width = "15em") %>% 
  row_spec(dim(results)[1], bold = TRUE)
```

### 3. What was the total paid amount for admissions for CKD and ESKD members across all years, by kidney disease stage?

```{r echo = FALSE, results = 'asis', message = FALSE}
grouped_counts <- all_data %>% 
  filter(kd_type != "None") %>% 
  group_by(kd_type) %>% 
  summarise(paid = sum(payment_amt, na.rm = TRUE))

sum_row <- lapply(grouped_counts,function(z) if (is.numeric(z)) sum(z) else '')
sum_row[1] <- "Total"
results <- rbind(grouped_counts,sum_row)
results$paid <- scales::dollar(results$paid)
colnames(results) <- c("Kidney Disease Stage","Paid Amount")

kable(results, align = 'lr', format.args = list(big.mark = ",")) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left") %>% 
  column_spec(1, width = "15em") %>% 
  row_spec(dim(results)[1], bold = TRUE)
```

### 4. What are the most common co-morbid chronic conditions for CKD and ESKD members?

```{r echo = FALSE, results = 'asis', message = FALSE}
comorbidities <- all_data %>% 
  filter(kd_type != "None") %>% 
  group_by(DESYNPUF_ID) %>% 
  summarise(
    has_alzheimer = max(has_alzheimer),
    has_heart_failure = max(has_heart_failure),
    has_chronic_kidney = max(has_chronic_kidney),
    has_cancer = max(has_cancer),
    has_copd = max(has_copd),
    has_depression = max(has_depression),
    has_diabetes = max(has_diabetes),
    has_ischemic_heart = max(has_ischemic_heart),
    has_osteoporosis = max(has_osteoporosis),
    has_arthritis = max(has_arthritis),
    has_stroke_tia = max(has_stroke_tia)
  ) %>% 
  summarise(
    comorbid_alzheimer = sum(has_alzheimer)/n(),
    comorbid_heart_failure = sum(has_heart_failure)/n(),
    #comorbid_chronic_kidney = sum(has_chronic_kidney)/n(),
    comorbid_cancer = sum(has_cancer)/n(),
    comorbid_copd = sum(has_copd)/n(),
    comorbid_depression = sum(has_depression)/n(),
    comorbid_diabetes = sum(has_diabetes)/n(),
    comorbid_ischemic_heart = sum(has_ischemic_heart)/n(),
    comorbid_osteoporosis = sum(has_osteoporosis)/n(),
    comorbid_arthritis = sum(has_arthritis)/n(),
    comorbid_stroke_tia = sum(has_stroke_tia)/n()
  )

results <- gather(comorbidities,
  "comorbid_alzheimer",
  "comorbid_heart_failure",
  #"comorbid_chronic_kidney",
  "comorbid_cancer",
  "comorbid_copd",
  "comorbid_depression",
  "comorbid_diabetes",
  "comorbid_ischemic_heart",
  "comorbid_osteoporosis",
  "comorbid_arthritis",
  "comorbid_stroke_tia",
  key = "Comorbidity",
  value = "Rate") 
results$Comorbidity <- c(
  "Alzheimer",
  "Heart Failure",
  #"Chronic Kidney Disease",
  "Cancer",
  "Chronic Obstructive Pulmonary Disease",
  "Depression",
  "Diabetes",
  "Ischemic Heart Disease",
  "Osteoporosis",
  "Rheumatoid Arthritis or Osteoarthritis",
  "Stroke/Transient Ischemic Attack"
)
results$Rate <- scales::percent(results$Rate, accuracy = 1)
results <- results %>% 
  arrange(desc(Rate))

kable(results, align = 'lr') %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left") %>% 
  column_spec(1, width = "25em")
```